Learning Multiple Networks via Supervised Tensor Decomposition

### 2020-10-02 tensor_regress/NeurIPSw/neuripsw_tensor.tex
workshop
revised abstract

\begin{abstract}
   We develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike unsupervised tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating optimization algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project. We identify the key global brain connectivity pattern and pinpoint the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas.
\end{abstract}

@@

In addition to the aforementioned challenges, many tensor datasets consist of non-Gaussian measurements. Classical tensor decomposition methods are based on minimizing the Frobenius norm of deviation, leading to suboptimal predictions for binary- or count-valued response variables. A number of supervised tensor methods have been proposed \citep{narita2012tensor, zhao2012higher, yu2016learning,lock2018supervised}. These methods often assume Gaussian distribution for the tensor entries, or impose random designs for the feature matrices, both of which are less suitable for applications of our interest. The gap between theory and practice means a great opportunity to modeling paradigms and better capture the complexity in tensor data.

@@

In classical tensor decomposition, tensor factorization is performed on either data tensor $\tY$ or mean tensor $\mathbb{E}(\tY)$. In the context of supervised tensor decomposition, we propose to factorize the latent parameter tensor $\tB$,
\begin{equation}\label{eq:factor}
\tB=\tC\times\{\mM_1,\ \ldots,\ \mM_K\},
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$. By the definition of multilinear rank, model equations~\eqref{eq:tensormodel} and~\eqref{eq:factor} imply the low-rankness $\mr=(r_1,\ldots,r_K)$ of the conditional mean tensor under the link function. We now reach our final model for supervised tensor decomposition,
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1, \ldots, \mX_K\mM_K\}\right),\notag\\
\text{with} \ &\ \mM_k^T\mM_k = \mI_{r_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\quad \text{for all }k=1,\ldots,K,
\end{align}
where the parameters of interest are $\mM_k$ and $\tC$. Note that model~\eqref{eq:decomp} assumes a fixed, known rank $\mr=(r_1,\ldots,r_K)$. Figure~\ref{fig:intro1}b provides a schematic illustration of our model. The features $\mX_k$ affect the distribution of tensor entries in $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. We call $\mX_k\mM_k$ the ``supervised tensor factors'' or ``sufficient features''~\citep{adragni2009sufficient}, and call $\mM_k$ the ``dimension reduction matrix.'' The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes, which links the conditional mean to the feature spaces, and thereby allows the identification of variations in the tensor data attributable to the side information. Our goal is to find $\mM_k$ and the corresponding $\tC$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

@@

>>>>

\begin{abstract}
   We develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike classical tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating optimization algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project. We identify the key global brain connectivity pattern and pinpoint the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas.
\end{abstract}

@@

In addition to the challenge of incorporating side information, many tensor datasets consist of non-Gaussian measurements. Classical tensor decomposition methods are based on minimizing the Frobenius norm of deviation, leading to suboptimal predictions for binary- or count-valued response variables. A number of supervised tensor methods have been proposed \citep{narita2012tensor, zhao2012higher, yu2016learning,lock2018supervised}. These methods often assume Gaussian distribution for the tensor entries, or impose random designs for the feature matrices, both of which are less suitable for applications of our interest. The gap between theory and practice means a great opportunity to modeling paradigms and better capture the complexity in tensor data.

@@

In classical tensor decomposition, tensor factorization is performed on either data tensor $\tY$ or mean tensor $\mathbb{E}(\tY)$. In the context of supervised tensor decomposition, we propose to factorize the latent parameter tensor $\tB$,
\begin{equation}\label{eq:factor}
\tB=\tC\times\{\mM_1,\ \ldots,\ \mM_K\},
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$. By the definition of multilinear rank, model equations~\eqref{eq:tensormodel} and~\eqref{eq:factor} imply the low-rankness $\mr=(r_1,\ldots,r_K)$ of the conditional mean tensor under the link function. We now reach our final model for supervised tensor decomposition,
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1, \ldots, \mX_K\mM_K\}\right),\notag\\
\text{with} \ &\ \mM_k^T\mM_k = \mI_{r_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\quad \text{for all }k=1,\ldots,K,
\end{align}
where the parameters of interest are $\mM_k$ and $\tC$. Note that model~\eqref{eq:decomp} assumes a fixed, known rank $\mr=(r_1,\ldots,r_K)$. Figure~\ref{fig:intro1}b provides a schematic illustration of our model. The features $\mX_k$ affect the distribution of tensor entries in $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. We call $\mX_k\mM_k$ the ``supervised tensor factors'' or ``sufficient features''~\citep{adragni2009sufficient}, and call $\mM_k$ the ``dimension reduction matrix.'' The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes, which links the conditional mean to the feature spaces, and thereby allows the identification of variations in the tensor data attributable to the side information. Our goal is to find $\mM_k$ and the corresponding $\tC$,thereby allowing us to reveal the relationship between side information $\mX$ and the observed tensor $\tY$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

@@

\newpage 
\section*{Broader Impact}

Our supervised tensor decomposition method is widely applicable to network analysis, dyadic data analysis, spatial-temporal model, and recommendation systems. We have shown the improved predictive power and enhanced interpretability by incorporating the interactive side information in tensor decomposition method.  The application to the brain connection dataset yields conclusions with sensible interpretations, suggesting the practical utility of the proposed approach.  We believe that our model enriches the research of tensor-based learning and is a powerful tool to boost scientific discoveries in various fields.

%We have developed a supervised tensor decomposition method with side information on multiple modes. One important challenge of tensor data analysis is the complex interdependence among tensor entries and between multiple features. Our approach incorporates side information as feature matrices in the conditional mean tensor. Our empirical results demonstrate the improved interpretability and accuracy over previous approaches. The application to the brain connection dataset yields conclusions with sensible interpretations, suggesting the practical utility of the proposed approach.  We believe that our model fills the gap between theory and practice and is a powerful tool to boost scientific discoveries.

%Our exponential family tensor regression  We have shown the improved predictive power and enhanced interpretability by incorporating interactions between multiple covariate matrices. We acknowledge the opportunities and challenges in the tensor regression modeling. We hope the method developed in this paper will open up new research directions towards richer tensor-based learning. All code and data are publicly available, and we encourage members of the machine learning community to participate in these exciting problems.
